---
title: "*Tetrahymena* p-hacking"
output: html_document
date: "2026-09-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will be needing the tidyverse
```{r}
library(tidyverse)
```

Step 1 of the masterplan is to load in the data. We are using the OurWorldInData COVID dataset and the OurWorldInData maize production dataset. Our "p-hacking" hypothesis will be that countries that produce more maize suffer from fewer COVID infections.

```{r}
covid_df <- read_csv("https://catalog.ourworldindata.org/garden/covid/latest/compact/compact.csv")

maize_df <- read_csv("https://ourworldindata.org/explorers/global-food.csv?v=1&csvType=full&useColumnShortNames=true&Food=Maize+%28corn%29&Metric=Production&Per+capita=false")

# Rename some of the columns to make our life easier later
names(maize_df)[1] <- "country"
names(maize_df)[length(names(maize_df))] <- "maize_prod"
```

For each country, we'll get average new cases per million people per day in 2021 and join this with maize production (in tonnes) from 1990.
```{r}
covid_df <-
  covid_df %>%
  filter(year(date) == 2021) %>%
  select(country, new_cases_per_million) %>%
  drop_na() %>%
  group_by(country) %>%
  summarise(total_new_cases_per_mil = sum(new_cases_per_million),
            recorded_days = n(),
            new_cases_per_day_mil = total_new_cases_per_mil/recorded_days)

maize_df <-
  maize_df %>%
  filter(year == 1990)

covid_df <-
  left_join(covid_df, maize_df, by=("country")) %>%
  select(country, new_cases_per_day_mil, maize_prod) %>%
  drop_na()
```

Let's see what this relationship looks like
```{r}
covid_df %>% ggplot(aes(maize_prod, new_cases_per_day_mil)) +
  geom_point()
```

That looks like it could use a log transformation on the x axis
```{r}
covid_df %>% ggplot(aes(maize_prod, new_cases_per_day_mil)) +
  geom_point() +
  scale_x_log10()
```

That's more like it. This looks like there's no relation, let's just confirm that
```{r}
check_lm <- lm(new_cases_per_day_mil ~ log(maize_prod), data=covid_df)
summary(check_lm)
```

Having confirmed there is no true relationship, we'll now engage in egregious statistical malpractice.

```{r}
# We'll simulate the p-hacking where you stop collecting data when you get a significant p
# Initialise our p_value from a model with 5 random countries
set.seed(5678)
covid_df_sample <- covid_df %>% slice_sample(n=5, replace=F)
remaining_countries <- covid_df %>% filter(!(country %in% covid_df_sample$country))

linmod <- lm(new_cases_per_day_mil ~ log(maize_prod), data=covid_df_sample)
p <- summary(linmod)$coefficients["log(maize_prod)", "Pr(>|t|)"]

i <- 0 # this will stop us getting trapped in the while loop forever

signif_thresh <- 0.05 # set alpha


while(p > signif_thresh & i < nrow(covid_df)){
  # Sample one new country
  sampled_country <- remaining_countries %>% slice_sample(n=1)
  covid_df_sample <- rbind(covid_df_sample, sampled_country)
  remaining_countries <- remaining_countries %>% filter(!(country %in% sampled_country$country))
  
  # Build a new linear model
  linmod <- lm(new_cases_per_day_mil ~ log(maize_prod), data=covid_df_sample)
  p <- summary(linmod)$coefficients["log(maize_prod)", "Pr(>|t|)"]
  i <- i+1
  print(p)
}
```

We got a significant p-value after a few iterations. Let's plot the data we sampled:
```{r}
summary(linmod)

covid_df_sample %>% ggplot(aes(maize_prod, new_cases_per_day_mil)) +
  geom_point() +
  geom_smooth(method="lm")
```

This looks kind of weird, let's commit more malpractice and remove the "outlier". Although you could argue that in a situation like this where you suspect your result is driven by one or a few outliers there is a case for seeing how robust your result is to their removal. But there are better ways to assess this e.g. Cook's distance in a linear regression like this.
```{r}
covid_df_sample_no_outliers <- covid_df_sample %>%
  filter(maize_prod < 3e+06)

linmod_no_outlier <- lm(new_cases_per_day_mil ~ log(maize_prod), data=covid_df_sample_no_outliers)
summary(linmod_no_outlier)
covid_df_sample_no_outliers %>% ggplot(aes(maize_prod, new_cases_per_day_mil)) +
  geom_point() +
  geom_smooth(method="lm")
```

Looks like that "outlier" was doing a lot for us. Let's commit more malpractice and put it back in now that we've seen the p-value from when we take it out.
```{r}
summary(linmod)

covid_df_sample %>% ggplot(aes(maize_prod, new_cases_per_day_mil)) +
  geom_point() +
  geom_smooth(method="lm")
```

To recap, the different forms of statistical malpractice we did were:

1. p-hacked by collecting data until we got a significant result (with significance threshold $\alpha = 0.05$)

2. The subsequent dataset had a weird-looking "outlier" which we removed because it looked a bit funny

3. That made our significant result disappear so we put it back in

A short summary of our findings could be something along the lines of:

Countries with higher maize production in 1990 had significantly lower COVID infection rates in 2021 (t=-2.254, p = 0.04206). We conclude consuming maize during childhood has a protective effect against adult COVID infection and call for further research into possible mechanisms underlying this effect.
